Sampling requirements to detect ecosystem change

### 2020-06-29 R/manuscript_draft.Rmd
modified citations

Number of references: 38<!--\total{citenum}-->

@@

Submission as: Article at \emph{Journal of Applied Ecology}

@@

Environmental monitoring is a key component of understanding and managing ecosystems. Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective. In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed. For example, sediment or ice cores can be quickly obtained in the field, but they require weeks or months of processing in a laboratory setting. Standard sub-sampling approaches often involve equally-spaced sampling. We use simulations to show how many samples, and which types of sampling approaches, are the most effective in detecting ecosystem change. We test these ideas with a case study of Cladocera community assemblage reconstructed from a sediment core. We demonstrate that standard approaches to sample processing are less efficient than an iterative approach. For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs. We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power. We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.

@@

Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a; @White2019e; @Lindenmayer2020a]. Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007]. Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a]. However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection [@Zhang2012]. Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed. Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lindenmayer2020a].

Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019e]. Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].

>>>>

Number of references: 40<!--\total{citenum}-->

@@

Submission as: Article at \emph{Ecological Indicators}

@@

Environmental monitoring is a key component of understanding and managing ecosystems. Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective. In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed. For example, sediment or ice cores can be quickly obtained in the field, but they require weeks or months of processing in a laboratory setting. Standard sub-sampling approaches often involve equally-spaced sampling. We use simulations to show how many samples, and which types of sampling approaches, are the most effective in detecting ecosystem change. We test these ideas with a case study of Cladocera community assemblage indicators reconstructed from a sediment core. We demonstrate that standard approaches to sample processing are less efficient than an iterative approach. For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs. We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power. We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.

@@

Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a; @White2019e; @Lindenmayer2020a]. Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007]. Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a]. However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection [@Zhang2012]. Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed. Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lengyel2018; @Lindenmayer2020a].

Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019e]. Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], imperfect detection [@Morant2020], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].
